High geomagnetic field intensity recorded by anorthosite xenoliths requires a vigorous late Mesoproterozoic geodynamo

### 2022-02-08 PNAS_manuscript/PNAS_manuscript.tex
Updates from Overleaf

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to provide constraints on the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, likely necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow \cite{Swanson-Hysell2021c}. However, strikingly low estimates of geomagnetic field strengths have been obtained ca. 565 Ma during the Ediacaran Period \cite{Bono2019a, Shcherbakova2019a,Thallner2021b} and ca. 370 Ma during the Devonian Period \cite{Shcherbakova2017a, Shcherbakova2021a, Hawkins2021a}, potentially indicating unusual periods of core dynamo regime at those times. The Ediacaran data have been interpreted to indicate that there was a progressively decaying field up to that time that was followed soon after by initial crystallization of the inner core \cite{Bono2019a}. However, the ability for the geomagnetic dipole field to reach similar low values in rather recent times when the inner core unambiguously exists (Fig. \ref{fig:PINT_compilation}) makes it challenging to unambiguously determine the timing of the initiation of inner core nucleation based on only a few records that span a relatively short time period.

@@

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=13 cm]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic images (A,B,C,D) and magnetic maps (E,F) of anorthosite from the Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column;  \citealp{Zhang2021b}) and a distinct anorthosite from the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given the complexities associated with more pronounced fabrics. Cross-polarized petrographic images of the Beaver River anorthosite (A,C) reveal plagioclase with a granoblastic texture with crystals that are largely free of large opaque inclusions. In contrast, plagioclase crystals in the Duluth Complex anorthosite have euhedral, interlocking crystals with an igneous foliation (B) and the plagioclase crystals contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [001] axis of the plagioclase. Magnetic field maps acquired with a quantum diamond microscope (QDM) on samples to which an IRM has been applied show that the relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite in (E) while large oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the overlapping magnetic fields above the oxides in (F). Magnetic field maps of (E) and (F) share the same spatial scale. Note the large contrast in color scale of panels E and F. B$_z$ denotes magnetic field intensity in the vertical direction of the field of view with positive values corresponding to the out of page direction. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

@@

These features of the Beaver River anorthosite xenoliths are distinct from many other anorthosites such as the anorthosite xenoliths within the Duluth Complex Anorthositic Series rocks---older intrusions within the Midcontinent Rift. To illustrate these difference, we present petrographic and magnetic imaging data from a sample of Duluth Complex anorthosite that was not targeted for paleointensity experiments. The plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display strong igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles that are typically aligned with the [001] axes of the plagioclase crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that these strongly magnetic needles within the plagioclase crystals often have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}D,F). Magnetic imaging also finds that orientations of magnetic needles within the Duluth plagioclase crystals depict the orientation of remanence magnetizations as is shown by the oblique acquired remanence to the direction of an applied IRM in Figure \ref{fig:Petro_QDM}F. Upon applying a second IRM of 300 mT in the orthogonal direction on the Duluth anorthosite, the applied field failed to rotate the remanence directions in plagioclase 1 but resulted in the reversal of remanence in plagioclase 2 which is still oblique to the applied field direction (Fig. S1). In contrast, the remanence magnetizations imaged in the Beaver River anorthosite xenoliths align parallel to the applied field directions upon applying a set of orthogonal external fields (Fig. \ref{fig:Petro_QDM}, S1). Titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex magma \cite{Miller1990a}.

@@

In addition to estimating paleointensity values by introducing a set of selection criteria to filter our experiment results, we apply an independent method of \citealp{Cych2021a} which uses all experimental data regardless of their NRM/TRM plot statistics to perform bias corrected estimation of paleointensity (BiCEP). This Bayesian probabilistic method is based on an assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Paterson2011a}. The posterior paleointensity distributions from these sites with high-quality specimen-level data are in agreement with the site-level averages developed using the selection criteria approach (Fig. \ref{fig:PINT_cooling_corrected}; Fig. S2). Overall, the high-quality results from the anorthosite xenoliths of the Beaver River diabase indicate that the anorthosite xenoliths record a high geomagnetic field ca. 1092 Ma.

@@

Rock magnetic data support that the anorthosite specimens that pass the paleointensity selection criteria have dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection and all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show both diabase and anorthosite contain (titano)magnetite as evidenced through the presence of the Verwey transition (Fig. S3;  \citealp{Verwey1939a, Feinberg2015a}). Anorthosite specimens from sites that yield successful paleointensity results have Verwey transition temperatures near 120 K as expected for stochiometric magnetite with minimal Ti \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have Verwey transitions that are suppressed toward lower temperatures (Fig. S3), indicating that magnetite grains in the specimens either have relatively higher Ti content or have been partially oxidized \cite{Ozdemir1993a}.

@@

We applied full TRMs to a set of anorthosite specimens with lab fields of 30, 50, 70, and 90 $\mu$T along specimen vertical axes. The results show that this set of anorthosite specimens did not saturate upon the applied fields and there is thus no need for correcting nonlinear thermal remanence acquisitions (\citealp{Selkin2007a}; Fig. S5).

@@

Another consequence of the anorthosite xenoliths having cooled in the interior of thick diabase intrusions is that slow cooling rates can bias paleointensity estimates toward higher values. Large differences in cooling rates between acquisition of an NRM in nature versus a TRM in the lab can result in overestimated paleointensities for single domain grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. From the thermal history model of ref.  \citealp{Zhang2021b}, we can estimate the duration over which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they fully blocked in their characteristic natural remanence magnetization ($\sim$500\textdegree C;  \citealp{Zhang2021b}). We find the cooling time to be on the order of $\sim$1 kyr, which corresponds to a cooling rate of $\sim1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster through the same temperature interval with an estimated cooling rate of $\sim1.3\times10^{-1}$ $^\circ$C s$^{-1}$. The significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of \citealp{Halgedahl1980a} (Fig. S4). This estimate on cooling rate effect is similar to the value of $\sim$35\% overestimate derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.74. The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level means are shown in Fig. \ref{fig:PINT_cooling_corrected}.

>>>>

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to provide constraints on the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, likely necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow \cite{Swanson-Hysell2021c}. However, strikingly low estimates of geomagnetic field strengths have been obtained ca. 565 Ma during the Ediacaran Period \cite{Bono2019a, Shcherbakova2019a,Thallner2021b} and ca. 370 Ma during the Devonian Period \cite{Shcherbakova2017a, Shcherbakova2021a, Hawkins2021a}, potentially indicating unusual periods of core dynamo regime at those times. The Ediacaran data have been interpreted to indicate that there was a progressively decaying field up to that time that was followed soon after by initial crystallization of the inner core \cite{Bono2019a}. However, the ability for the geomagnetic dipole field to reach similar low values in rather recent times when the inner core unambiguously exists (Fig. \ref{fig:PINT_compilation}) makes it challenging to unambiguously settle the timing of the initiation of inner core nucleation based on such few records that span a relatively short time period.

@@

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=13 cm]{Petro_QDM_full.png}
\caption{\footnotesize{Thin section petrographic images (A,B,C,D) and magnetic maps (E,F) of anorthosite from the Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column;  \citealp{Zhang2021b}) and a distinct anorthosite from the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given the complexities associated with more pronounced fabrics. Cross-polarized petrographic images of the Beaver River anorthosite (A,C) reveal plagioclase with a granoblastic texture with crystals that are largely free of large opaque inclusions. In contrast, plagioclase crystals in the Duluth Complex anorthosite have euhedral, interlocking crystals with an igneous foliation (B) and the plagioclase crystals contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [001] axis of the plagioclase. Magnetic field maps acquired with a quantum diamond microscope (QDM) on samples to which an IRM has been applied show that the relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite in (E) while large oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the overlapping magnetic fields above the oxides in (F). (E)-(H) show an experiment performed on both samples where we apply a first field along the y axis of the field of view and then apply a second field of 300 mT orthogonal to the first field direction. The isothermal remanent magnetizations acquired by both anorthosites were mapped after the application of each field. The magnetic images show that dipole-like remanent magnetizations of Beaver River anorthosite align well with the first applied field direction (E) and rotated to align with the second applied field direction with little anisotropic behavior (G). In contrast, the remanence magnetizations carried by the magnetic needles in plagioclase 1 of the Duluth anorthosite xenolith align well with the first applied field but those in plagioclase 2 tend to acquire an oblique remanence direction with respect to the field direction (F). After the application of a 300 mT external field, magnetization of those needles in plagioclase 1 did not reverse due to strong shape anisotropy whereas those needles in plagioclase 2 reversed but still acquired remanence oblique to the field direction (H). This experiment illustrates that the Beaver River anorthosite xenolith has much lower degrees of magnetic anisotropy than the Duluth anorthosite xenolith. B$_z$ denotes magnetic field intensity in the vertical direction with positive values corresponding to the out of page direction. Note the large contrast in magnetic field scale of (E) and (G) with those in (F) and (G). The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

@@

These features of the Beaver River anorthosite xenoliths are distinct from many other anorthosites such as the anorthosite xenoliths within the Duluth Complex Anorthositic Series rocks---older intrusions within the Midcontinent Rift. To illustrate these difference, we present petrographic and magnetic imaging data from a sample of Duluth Complex anorthosite that was not targeted for paleointensity experiments. The plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display strong igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles that are typically aligned with the [001] axes of the plagioclase crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that these strongly magnetic needles within the plagioclase crystals often have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}D,F). Magnetic imaging also finds that orientations of magnetic needles within the Duluth plagioclase crystals depict the orientation of remanence magnetizations as is shown by the oblique acquired remanence to the direction of an applied IRM in Figure \ref{fig:Petro_QDM}F. Upon applying a second IRM of 300 mT in the orthogonal direction on the Duluth anorthosite, the applied field failed to rotate the remanence directions in plagioclase 1 but resulted in the reversal of remanence in plagioclase 2 which is still oblique to the applied field direction (Fig. \ref{fig:Petro_QDM}). In contrast, the remanence magnetizations imaged in the Beaver River anorthosite xenoliths align parallel to the applied field directions upon applying a set of orthogonal external fields (Fig. \ref{fig:Petro_QDM}). Titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex magma \cite{Miller1990a}.

@@

In addition to estimating paleointensity values by introducing a set of selection criteria to filter our experiment results, we apply an independent method of \citealp{Cych2021a} which uses all experimental data regardless of their NRM/TRM plot statistics to perform bias corrected estimation of paleointensity (BiCEP). This Bayesian probabilistic method is based on an assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Paterson2011a}. The posterior paleointensity distributions from these sites with high-quality specimen-level data are in agreement with the site-level averages developed using the selection criteria approach (Fig. \ref{fig:PINT_cooling_corrected}; Fig. S1). Overall, the high-quality results from the anorthosite xenoliths of the Beaver River diabase indicate that the anorthosite xenoliths record a high geomagnetic field ca. 1092 Ma.

@@

Rock magnetic data support that the anorthosite specimens that pass the paleointensity selection criteria have dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection and all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show both diabase and anorthosite contain (titano)magnetite as evidenced through the presence of the Verwey transition (Fig. S2;  \citealp{Verwey1939a, Feinberg2015a}). Anorthosite specimens from sites that yield successful paleointensity results have Verwey transition temperatures near 120 K as expected for stochiometric magnetite with minimal Ti \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have Verwey transitions that are suppressed toward lower temperatures (Fig. S2), indicating that magnetite grains in the specimens either have relatively higher Ti content or have been partially oxidized \cite{Ozdemir1993a}.

@@

We applied full TRMs to a set of anorthosite specimens with lab fields of 30, 50, 70, and 90 $\mu$T along specimen vertical axes. The results show that this set of anorthosite specimens did not saturate upon the applied fields and there is thus no need for correcting nonlinear thermal remanence acquisitions (\citealp{Selkin2007a}; Fig. S4).

@@

Another consequence of the anorthosite xenoliths having cooled in the interior of thick diabase intrusions is that slow cooling rates can bias paleointensity estimates toward higher values. Large differences in cooling rates between acquisition of an NRM in nature versus a TRM in the lab can result in overestimated paleointensities for single domain grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. From the thermal history model of ref.  \citealp{Zhang2021b}, we can estimate the duration over which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they fully blocked in their characteristic natural remanence magnetization ($\sim$500\textdegree C;  \citealp{Zhang2021b}). We find the cooling time to be on the order of $\sim$1 kyr, which corresponds to a cooling rate of $\sim1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster through the same temperature interval with an estimated cooling rate of $\sim1.3\times10^{-1}$ $^\circ$C s$^{-1}$. The significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of \citealp{Halgedahl1980a} (Fig. S3). This estimate on cooling rate effect is similar to the value of $\sim$35\% overestimate derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.74. The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level means are shown in Fig. \ref{fig:PINT_cooling_corrected}.

### 2022-02-08 PNAS_manuscript/Supporting Info.tex
Updates from Overleaf

\begin{figure*}[h!]
\noindent\includegraphics[width=0.8\textwidth]{Petro_QDM_SI.png}
\centering
\caption{{Petrographic images and magnetic images from the main manuscript with additional anisotropy experiment results. In A-F, we show that the Beaver River anorthosite often lack large oxide needles within plagioclase crystals carrying strong magnetic remanence  differences of magnetic carriers which is in contrast to what is seen in Duluth Complex anorthosites. E-H show an experiment performed on both anorthosite xenoliths where we apply a first field along the y axis of the field of view and then apply a second field of 300 mT orthogonal to the first field direction. The isothermal remanent magnetizations acquired by both anorthosites were imaged with a quantum diamond microscope after the application of each field. The magnetic images show that dipole-like remanent magnetizations of Beaver River anorthosite align well with the first applied field direction and rotated to align with the second applied field direction with little anisotropic behavior. In contrast, the remanence magnetizations carried by the magnetic needles in plagioclase 1 of the Duluth anorthosite xenolith align well with the first applied field but those in plagioclase 2 tend to acquire an oblique remanence direction with respect to the field direction (F). After the application of a 300 mT external field, magnetization of those needles in plagioclase 1 did not reverse due to strong shape anisotropy whereas those needles in plagioclase 2 reversed but still acquired remanence oblique to the field direction. This experiments illustrates that the Beaver River anorthosite xenolith has much lower degrees of magnetic anisotropy than the Duluth anorthosite xenolith. B$_z$ denotes magnetic field intensity in the vertical direction with positive values corresponding to the out of page direction.  }}
\label{fig:Petro_QDM_SI}
\end{figure*}
\clearpage

>>>>

% \begin{figure*}[h!]
% \noindent\includegraphics[width=0.8\textwidth]{Petro_QDM_SI.png}
% \centering
% \caption{{Petrographic images and magnetic images from the main manuscript with additional anisotropy experiment results. In A-F, we show that the Beaver River anorthosite often lack large oxide needles within plagioclase crystals carrying strong magnetic remanence  differences of magnetic carriers which is in contrast to what is seen in Duluth Complex anorthosites. E-H show an experiment performed on both anorthosite xenoliths where we apply a first field along the y axis of the field of view and then apply a second field of 300 mT orthogonal to the first field direction. The isothermal remanent magnetizations acquired by both anorthosites were imaged with a quantum diamond microscope after the application of each field. The magnetic images show that dipole-like remanent magnetizations of Beaver River anorthosite align well with the first applied field direction and rotated to align with the second applied field direction with little anisotropic behavior. In contrast, the remanence magnetizations carried by the magnetic needles in plagioclase 1 of the Duluth anorthosite xenolith align well with the first applied field but those in plagioclase 2 tend to acquire an oblique remanence direction with respect to the field direction (F). After the application of a 300 mT external field, magnetization of those needles in plagioclase 1 did not reverse due to strong shape anisotropy whereas those needles in plagioclase 2 reversed but still acquired remanence oblique to the field direction. This experiments illustrates that the Beaver River anorthosite xenolith has much lower degrees of magnetic anisotropy than the Duluth anorthosite xenolith. B$_z$ denotes magnetic field intensity in the vertical direction with positive values corresponding to the out of page direction.  }}
% \label{fig:Petro_QDM_SI}
% \end{figure*}
% \clearpage
